---
title: "HOMEWORK_6"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)
library(scales)
```

Problem 1¶
10 points

For each of the following functions, provide a <100 character description (in your own words) and a URL reference.

! - checks to see if 2 things are NOT equal (not equal = true) (itsalocke.com)
is.na - determines true/false for missing values (statmethods.net)
is.numeric - checks to see if an object is a number (geeksforgeeks.org)
anti_join - returns unmatched rows from one one dataframe compared to the other (sparkbyexamples.com)
desc - Sorts a vector in descending order (dplyr.tidyverse.org)
dplyr::slice - manipulate rows in a dataframe by location in the table (dplyr.tidyverse.org)
dplyr::across - Do a funtion to multiple specific columns in a dataframe (statology.org)


Problem 2
10 points

Add a comment above each code line below explaining what the code line does and/or why that code line is necessary.

Keep each comment to less than 2 lines per line of code and < 80 chars per line

```{r}
# read in a tsv file and call it "annotations"
annotations <- read_tsv("ftp://ftp.ebi.ac.uk/pub/databases/genenames/new/tsv/locus_groups/protein-coding_gene.txt") %>%
    # choose which columns to work with (exclude the others)
    select(ensembl_gene_id, symbol, name, gene_group, ccds_id) %>%
    # subset the data that is not NA in the ccds_id column 
    filter(!is.na(ccds_id)) %>%
  # print the resulting dataframe  
  print()
```


```{r}
# read in another tsv and call it "data"
data <- read_tsv("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE89nnn/GSE89183/suppl/GSE89183_Counts.txt.gz") %>%
  # change the name of the column called "ENSEMBL gene" to "ensembl_gene_id"
  rename(ensembl_gene_id = `ENSEMBL gene`) %>%
  # print the resulting dataframe
  print()
```


Problem 3
10 points

Using the code below:

Convert both axes to log10 instead of linear scales.
Show axis tick labels as 10^0^, 10^1^, 10^2^,10^3^, 10^4^, 10^5^ for both axes.
There are too many points overlapping in certain regions. Use a different geom_ function to convey to your reader how many overlapping points are present in each region.


```{r}
data %>%
  select(CD34_shRPL5_RNA_1, CD34_shRPS19_RNA_1) %>%
  ggplot(aes(x = CD34_shRPL5_RNA_1, y = CD34_shRPS19_RNA_1)) +
  geom_count(alpha = 0.25) + scale_y_continuous(trans='log10',
                     breaks=trans_breaks('log10', function(x) 10^x),
                     labels=trans_format('log10', math_format(10^.x))) +
                scale_x_continuous(trans='log10',
                     breaks=trans_breaks('log10', function(x) 10^x),
                     labels=trans_format('log10', math_format(10^.x)))
```
Problem 4
10 points

Write a code cell to select the following columns from the data variable you created above and reassign back to data.

Columns to select: ensembl_gene_id, columns that have either RPL5, RPS19, or Luc in their name.

Instead of listing every column, use matches and a succinct regular expression to select the above columns.

```{r}
data <- data %>% select(ensembl_gene_id, matches("RPL5"), matches("RPS19"), matches("Luc"))
```

```{r}
colnames(data)
```

Problem 5
10 points

Write a code cell to filter the result from Problem 4 to include only rows where each of the 12 numerical columns you selected has 50 counts or more and reassign back to data. This is a simple way to avoid genes that have very low counts.

You might be tempted to do this step separately for each of the 12 columns, but instead use the across function you learned above or the pivot_longer function from the class (either way is fine).


```{r}
data <- data %>% filter(across(c(2:13), ~. >=50))
```

Problem 6
10 points

Write a code cell to divide each of the 12 numerical columns by the corresponding median value for each column and reassign back to data. This median normalization is typically done in high-throughput experiments after filtering to normalize for sample-to-sample difference in read depth.

Again, you can write lot less code if you use the mutate and across function you learned above.


```{r}
data <- data %>% mutate_if(is.numeric, funs(./median(.)))
```


After we do the above filtering and median-normalization, let us calculate translation efficiency as the average ratio of the RPF and RNA reads for each treatment condition. Then we calculate how this translation efficiency changes between target (rpl5 and rps19) and control (luc) shRNAs.

The code implementing the above steps is shown below, but it has a few errors. Correct them.

```{r}
lfc <- data %>%
  mutate(mean_rpl5_te = ((CD34_shRPL5_RPF_1 + CD34_shRPL5_RPF_2) /
                            (CD34_shRPL5_RNA_1 + CD34_shRPL5_RNA_2))) %>%
  mutate(mean_rps19_te = ((CD34_shRPS19_RPF_1 + CD34_shRPS19_RPF_2) /
                            (CD34_shRPS19_RNA_1 + CD34_shRPS19_RNA_2))) %>%
  mutate(mean_shluc_te = ((CD34_shLuc_RPF_1 + CD34_shLuc_RPF_2) /
                            (CD34_shLuc_RNA_1 + CD34_shLuc_RNA_2))) %>%
  select(ensembl_gene_id, mean_rpl5_te, mean_rps19_te, mean_shluc_te) %>%
  mutate(lfc_te_rpl5 = log2(mean_rpl5_te / mean_shluc_te),
         lfc_te_rps19 = log2(mean_rps19_te / mean_shluc_te))
```

Problem 8
10 points

Write code that will create a new dataframe called mean_lfc from lfc containing a new column called avg_lfc. avg_lfc should be the average of the log2 fold-change in TE (lfc_te) upon knockdown of RPL5 and RPS19.

Then select only the gene id column and the new column that you just created (this will be your new dataframe mean_lfc).


```{r}

mean_lfc <- lfc %>% 
  mutate (avg_lfc = ((lfc_te_rpl5 + lfc_te_rps19)/2)) %>%
  select(ensembl_gene_id, avg_lfc)


```

10 points

Write code to join the mean_lfc dataframe with the annotations dataframe created at the top of the document and assign back to mean_lfc


```{r}
mean_lfc <- mean_lfc %>%
  inner_join(annotations, by = "ensembl_gene_id")
```

Problem 10
10 points

Write code to select only the bottom 10 genes with the lowest avg_lfc and display the gene symbol, gene name and avg_lfc for these genes.
Create a figure using ggplot2 to visualize these results.
Write a few sentences to justify the choices you made when creating your figure.


```{r}
mean_lfc <- mean_lfc %>% arrange(avg_lfc)

top10 <- mean_lfc %>% slice (1:10)

print(top10)
```


```{r}
ggplot(top10, aes(x=symbol, y=avg_lfc)) +
  geom_col(color = "black", fill="blue") + ggtitle("Plot of average LFC by gene (bottom 10 values)") +
  xlab("Gene") + ylab("Average LFC") + theme(plot.title = element_text(hjust = 0.5))
```
Since I do not have any context for the data in this exercise, I plotted the average LFC values vs. the gene names "as is" - If I knew more about the data, I might have plotted it differently. I used geom_col instead of geom_bar so that I could define the data on the y-axis.  I made the bars blue with a black outline because it's pretty. I also centered the title to make it even prettier.

Credits: I worked with Sophie Kogut and Theresa Chen on this homework. Jacob was helpful. Here are some websites I visited while I was working:
https://dplyr.tidyverse.org/reference/across.html
https://www.statology.org/ggplot2-log-scale/
https://www.datasciencemadesimple.com/join-in-r-merge-in-r/
https://bookdown.org/aschmi11/RESMHandbook/data-visualization-with-ggplot.html
